---
title: "Greenhouse Gas Emissions: Comparing the World's Highest and Lowest Contributors"
subtitle: "SDS 164A Fall 2024"
output:
  html_document:
    theme: readable
    highlight: breezedark
author: Grace Nichols, Shen Rothermel
editor_options: 
  chunk_output_type: console
---

```{r Setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center", out.width = "110%")

library(tidyverse)
library(ggthemes)
library(viridis)

# Because one of us has a package with a select function which conflicts with dplyr's function
select <- dplyr::select
```

## Introduction

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Our question is as follows: among six countries included in our data set, three with some of the lowest emissions recorded and three with some of the highest emissions, how do their respective greenhouse gas emissions compare and contrast? The following analysis will help us to unpack the intricacies of a critical global issue: climate change. Humans are a driving force of the climate crisis[^1] and, in order to craft an effective array of possible solutions, we need first to assess where emissions are stemming from. Knowing the sources of emissions, such as by industry sector and geographic location, provides us with necessary information we can use when attempting to mitigate the worst impacts of climate change. This analysis is essential because it is not fair to treat all countries and industries the same, given their noticeable differences in polluting levels. To construct equitable climate policies across the globe, one must know where the majority and the minority of greenhouse gasses are being emitted.

[^1]: “The Causes of Climate Change.” NASA Science, NASA, 23 Oct. 2024, <https://science.nasa.gov/climate-change/causes/>.

## Data

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Our data comes from the Statistics Division of the Department of Economic and Social Affairs, in the United Nations. Various countries have independently, with assistance from the UN, gathered national data on CO2 and greenhouse gas emissions and reported this to the UN, which compiles it all into publicly available databases. This data has been reported and compiled annually from 2000 to 2021, including 1990, though many countries have not been consistent. Thus, there is some missing annual data, and some countries’ GHG emission levels data have not been recently updated. We will be looking at tonnes of CO2 produced (thousands or millions, depending on the data set) as well as millions of tonnes of GHG produced in total. For CO2, we will be utilizing available annual data to look at trends in levels of CO2 emissions over the past three decades. For GHG, we only have data from the most recent annual data for each country, which will be used primarily for comparing emission levels between countries. Throughout our graphs, we choose to focus on the following countries: Monaco, Malta, Iceland, Japan, the Russian Federation, and the United States of America. The first three have relatively low emission levels across the board, while the last three stand opposite with very high emission levels.

```{r Load Data, include = FALSE}
# Load in data and clean

# GHG Emissions
GHG_Emissions <-
  read_csv("GHG_Emissions.csv",
           col_names = c("CountryID", # Rename variables
                         "Country",
                         "Latest_Year_Available",
                         "Total_GHG_Emissions_Exclude_LULUCF_mio",
                         "Percent_Change_Since_1990",
                         "GHG_Per_Capita",
                         "Total_GHG_Emissions_Include_LULUCF_mio"),
           col_types = "nfndddd", # Set variable datatypes
           na = c("...", ""), # Interpret invalid values
           skip = 2) # Skip first two rows (verbose metadata)

# GHG Emissions Sector
GHG_Emissions_Sector <-
  read_csv("GHG_Emissions_by_Sector.csv",
           col_names = c("CountryID", # Rename variables
                         "Country",
                         "Latest_Year_Available",
                         "Total_GHG_Emissions_Exclude_LULUCF_thou",
                         "GHG_Energy_thou",
                         "GHG_Energy_Transport_thou",
                         "GHG_Industrial_thou",
                         "GHG_Agriculture_thou",
                         "GHG_Waste_thou",
                         "GHG_LULUCF_thou"),
           col_types = "nfnddddddd", # Set variable datatypes
           col_select = !(11:17), # Remove unnecessary columns
           na = c("...", ""), # Interpret invalid values
           skip = 2) # Skip first two rows (verbose metadata)

# CO2 Emissions Time Series
CO2_Emissions_Time_Series <-
  read_csv("CO2_Emissions_time series.csv",
           col_types = "nfnnnnnnnnnnnnnnnnnnnnnnn", # Set variable datatypes
           na = c("...", "")) # Interpret invalid values

# Pivot the Time Series Dataset
CO2_Emissions_Time_Series_Tidy <-
  CO2_Emissions_Time_Series |> 
  pivot_longer(`1990`:`2021`, # Pivot each year variable to a value under a single variable
               names_to = "Year",
               values_to = "CO2_Emissions_thous", # Values under Year are CO2 emissions
               values_drop_na = T) # Drop NAs

# CO2 Emissions
CO2_Emissions <-
  read_csv("CO2_Emissions.csv",
           col_names = c("CountryID", # Rename variables
                         "Country",
                         "Latest_Year_Available",
                         "CO2_Emissions_mio",
                         "Percent_Change_Since_1990",
                         "CO2_Per_Capita"),
           col_types = "nfnddd", # Set variable datatypes
           na = c("...", ""), # Interpret invalid values
           skip = 2) # Skip first two rows (verbose metadata)

# Countries to focus on for visualizations
focus_countries <- c("Monaco", # These three are very low in overall emissions
                     "Malta",
                     "Iceland",
                     "United States of America", # These three are very high in overall emissions
                     "Russian Federation",
                     "Japan")
```

### Carbon Dioxide Emissions

```{r CO2 Emissions Levels between Countries, fig.cap = "Figure 1. In terms of CO2 emissions levels, the United States of America leads ahead of the Russian Federation, followed closely by Japan. Monaco is the lowest out of the six countries in CO2 emissions levels, with Malta next and then Iceland. Gray lines represent all other countries included in the data set. The six highlighted countries are those with the lowest and highest CO2 emissions among countries with data for all years spanning from 2000 to 2021. CO2 emissions are measured in orders of magnitude or powers of 10. Data was recorded and compiled by the United Nations from countries around the world over the course of three decades. Year of data is reported, along with the country's emissions levels measured in orders of magnitude of tonnes of CO2 for that year.", fig.alt = "This scatterplot shows the trend of CO2 emissions levels between six chosen countries throughout two decades. The x-axis is the year from 2000 to 2021 and the y-axis is CO2 emissions levels in orders of magnitude of tonnes of CO2 equivalent. For the entire time range, the United States of America has the highest CO2 emissions out of the six countries at 9.75 orders of magnitude, followed by the Russian Federation and then closely by Japan. After a large jump over 3 orders of magnitude, next is Iceland, Malta, and lastly Monaco with under 5 orders of magnitude in tonnes of CO2 emissions."}
# Minor alterations to time series dataset for easier graphing
co2etst2 <- CO2_Emissions_Time_Series_Tidy |> 
  filter(Year != 1990) |> # Exclude year 1990 observations, since series technically starts in 2000
  mutate(Year = as.numeric(Year), # Convert Year to continuous
         # Convert CO2 Emissions from thousands of tonnes to just tonnes, then log10 it
         # This gives us orders of magnitude to make it easier
         # to compare countries across a very large range
         CO2_Emissions_thous_log = log10(CO2_Emissions_thous * 1000),
         # Factor reorder Country by CO2 emission levels for intuitive visualization
         Country = fct_reorder2(Country, Year, CO2_Emissions_thous_log))

# Same dataset but with just focus countries
co2etst2_focus <- co2etst2 |> 
  filter(Country %in% focus_countries)

# Visualization for CO2 Emissions Levels between Countries
co2etst2 |> 
  ggplot(aes(x = Year,
             y = CO2_Emissions_thous_log,
             group = Country)) +
  # Lines for all countries, goes in "background"
  geom_smooth(color = "gray", linewidth = 0.2, se = F, show.legend = F) +
  # Points for focus countries
  geom_point(data = co2etst2_focus,
             # Shape & color by country to more easily distinguish them
             aes(color = Country)) +
  # Lines for focus countries
  geom_smooth(data = co2etst2_focus,
              # Color by country to more easily distinguish them
              aes(color = Country),
              linewidth = 0.5, se = F, show.legend = F) +
  scale_x_continuous(guide = guide_axis(angle = 30)) + # Tilt the x-axis labels
  # Make the y-axis labels' meaning a little more intuitive
  scale_y_continuous(breaks = c(4, 5, 6, 7, 8, 9, 10),
                     labels = parse(text = c("10^4", "10^5", "10^6", "10^7", "10^8", "10^9", "10^10"))) + 
  labs(title = "CO2 Emissions Levels between Countries",
       subtitle = "2000 - 2021",
       y = "CO2 Emissions in Orders of Magnitude") +
  theme_dark() +
  scale_color_viridis_d(option = "magma")
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Although this graph cannot completely help us answer the question posed at the beginning, it provides insight into the stark differences in CO2 emissions among the six countries of interest. It seems that the United States emits the most carbon dioxide, Monaco emits the lowest amount of CO2, and the other countries lie in between the two. We have yet to see how emissions have changed from 2000 to 2021 for each individual country or each country’s total greenhouse gas emissions. However, the information provided by the graph strongly suggests that policy makers will need to focus more of their energy on implementing policies that reduce CO2 being emitted by the United States and other high polluting countries such as the Russian Federation.

```{r Percentage CO2 Emissions Changes by Country, fig.cap = "Figure 2. Monaco, Malta, Japan, and the United States of America trended down in CO2 emission levels (in millions of tonnes) between 2000 and 2021, while Iceland and the Russian Federation have had an increase during that time range. Data was recorded and compiled by the United Nations from countries around the world over three decades. Year of data is reported, along with the country's emissions levels measured in millions of tonnes of CO2 for that year.", fig.alt = "This faceted scatter and line plot shows trends in CO2 emissions levels for each country over two decades. The x-axis is the year from 2000 to 2021 and the y-axis is percent change in CO2 emmissions in that time, ranging from ranging from approximately negative 50% to positive 30%. A downward smooth curve for Monaco, Malta, Japan, and the United States of America suggest a negative trend in CO2 emissions over the two decades for each of them. Conversely, an upward smooth curve for Iceland and the Russian Federation during this time period suggests a positive trend  in pecent change in CO2 emissions."}
# Visualization for percentage changes in CO2 emissions levels from 2000 to 2021 by country
CO2_Emissions_Time_Series |> 
  filter(Country %in% focus_countries) |> # Pick out only observations for focus countries
  # Recalulate values under each year in terms of percent change, using 2000 as the base case
  mutate(across(`2001`:`2021`, \(x) (x / `2000` - 1) * 100),
         `2000` = 0) |> 
  # Pivot longer to get tidy data
  pivot_longer(cols = `2000`:`2021`,
               names_to = "Year",
               values_to = "percent_change_CO2_emissions") |> 
  mutate(Year = as.numeric(Year), # Convert Year to continuous
         # Factor reorder Country by percent change for intuitive visualization
         Country = fct_reorder2(Country, Year, percent_change_CO2_emissions)) |> 
  ggplot(aes(x = Year,
             y = percent_change_CO2_emissions,
             color = Country)) +
  geom_line(linewidth = 1, linejoin = "bevel") +
  scale_y_continuous(n.breaks = 8,
                     # Append a percent symbol to the y-axis break labels
                     labels = \(x) paste(x, "%", sep = "")) + 
  labs(title = "Percent Change in CO2 Emissions Levels Since 2000 by Country",
       subtitle = "2000 - 2021",
       x = "Year",
       y = "Percent Change in CO2 Emissions Since 2000") +
  theme_dark() +
  scale_color_viridis_d(option = "magma")
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The graph above provides a partial answer for our question. Unsurprisingly, through the past 20 years, CO2 emissions have changed for all countries under investigation. From 2000 to 2021, Monaco’s CO2 emissions have steadily declined. The CO2 emissions of other countries such as Malta, the United States of America, and Japan have begun to invariably decline sometime after the year 2000. Unlike the other countries mentioned, the Russian Federation and Iceland have been increasing their carbon dioxide emissions since 2000. As large polluters, the direction in which the United States and Japan are moving regarding CO2 emissions would appear to be favorable. Even so, it is important to remember from the previous graph that Japan and the United States are still polluting in billions of tonnes while Malta and Iceland are polluting in millions of tonnes and Monaco is polluting in fractions of that. Thus, the United States and Japan will need to continue to reduce their emissions, possibly even more aggressively, in order to keep global degrees of warming at a desired amount [^2] [^3].

[^2]: Fendt, Lindsay. “Why Did the IPCC Choose 2° C as the Goal for Limiting Global Warming?” MIT Climate Portal, 22 June 2021, <https://climate.mit.edu/ask-mit/why-did-ipcc-choose-2deg-c-goal-limiting-global-warming>.

[^3]: “A Degree of Concern: Why Global Temperatures Matter.” NASA Science, NASA, 19 June 2019, <https://science.nasa.gov/earth/climate-change/vital-signs/a-degree-of-concern-why-global-temperatures-matter/>.

### Greenhouse Gas Emissions

```{r Per Captia GHG Emissions by Country, out.width = "80%", fig.cap = "Figure 3. In 2021, the United States led ahead in per capita GHG emmisisons followed by the Russian Fderation, Iceland, Japan, Malta, and lastly Monaco. Data was compiled by the United Nations from countries around the world according to the latest year for recorded data available. The country's per capita Greenhouse gas emissions are measured in tonnes of CO2 equivalent for that year.", fig.alt = "This is a bar chart with the countries Monaco, Malta, Japan, Iceland, the Russian Federation, and the United States of America, in this order, on the x-axis. Each bar's height correlates to the country's greenhouse gas emissions per capita in tonnes of CO2 equivalent, ranging from about 4 up to just under 20. Given the ordering of the countries along the x-axis, it is clear that the USA has the greatest per capita emissions, with Monaco sitting on the opposite end."}
# Visualization for GHG emissions per capita by country
GHG_Emissions |> 
  filter(Country %in% focus_countries) |> # Pick out only observations for focus countries
  mutate(Country = fct_reorder(Country, GHG_Per_Capita)) |> 
  ggplot(aes(x = Country, # Arrange the columns in ascending order
             y = GHG_Per_Capita,
             fill = Country)) +
  geom_col(show.legend = F) +
  scale_x_discrete(guide = guide_axis(angle = 30)) +
  labs(title = "Per Capita GHG Emissions by Country",
       subtitle = "2021",
       x = "Country",
       y = "GHG Emissions Per Capita (tonnes of CO2 equivalent)") +
  theme_dark() +
  scale_fill_viridis_d(option = "magma")
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;This graph gives us insight into the question posed at the beginning by showing us what each country's greenhouse gas emissions per capita were - how much, on average, the country produced for each person. Even at this level, the US holds a significant lead over the other countries at about 18 tonnes of CO2 equivalent per person.

```{r Sector Preparation, include = F}
# Sector data tidied up for easier graphing
tidy_sector <- GHG_Emissions_Sector |> 
  filter(Country %in% focus_countries) |> # Pick out only observations for focus countries
  mutate(GHG_Energy_thou = GHG_Energy_thou - GHG_Energy_Transport_thou) |> # Separate transport numbers from energy numbers
  pivot_longer(cols = 5:10,
               names_to = "Sector",
               values_to = "GHG_Emissions_thou") |> 
  # Negative values indicate sequestration, which is opposite from emissions, our main interest
  mutate(GHG_Emissions_thou = ifelse(is.na(GHG_Emissions_thou) | GHG_Emissions_thou < 0, 0, GHG_Emissions_thou),
         # Change sector names to be more intuitive
         Sector = fct_recode(Sector,
                             "Agriculture" = "GHG_Agriculture_thou",
                             "Energy" = "GHG_Energy_thou",
                             "Transport" = "GHG_Energy_Transport_thou",
                             "Industrial" = "GHG_Industrial_thou",
                             "Land Use" = "GHG_LULUCF_thou",
                             "Waste" = "GHG_Waste_thou"),
         Country = fct_reorder(Country, GHG_Emissions_thou))
```

```{r GHG Emissions Proportion Levels by Sector, fig.cap = "Figure 4. In 2021, proportions in greenhouse gas emissions for each sector were relatively similar for Monaco, Japan, the Russian Federation, and the USA, but Monaco has a noticely smaller proportion in energy and Iceland has a significantly larger proportion in land use. Data was compiled by the United Nations from countries around the world according to the latest year for recorded data available. Each proportion is calculated as the ratio of the country's sector emissions to the country's total emissions for that year. Negative emissions for land use from Japan, Monaco, the Russian Federation, and the USA were removed, as they represent sequestration. Monaco lacks data for its agriculture data for this year.", fig.alt = "This is a segmented bar chart with the countries Monaco, Malta, Iceland, Japan, the Russian Federation, and the USA on the x-axis. Each bar is segmented into six sections: agriculture, energy, transport, industrial, land use, and waste. Each segment's size corresponds to its proportion out of the country's total greenhouse gas emissions. Nearly all of the countries have similar proportions across all the sectors, with energy being at least 45% and as much as 70%. However, Iceland is an exception, with land use being its dominating sector at over 60%."}
tidy_sector |> 
  ggplot() +
  # Stack sector data on each other to have a single column for each country
  geom_col(aes(x = Country,
               y = GHG_Emissions_thou,
               fill = Sector),
           position = "fill") +
  scale_x_discrete(guide = guide_axis(angle = 30)) +
  labs(title = "Greenhouse Gas Emission Proportion Levels by Sector",
       subtitle = "2021",
       y = "Percent of GHG Emissions",
       caption = "No data for Monaco's agriculture sector, sequestration data (negative emissions) was removed") +
  theme_dark() +
  scale_fill_viridis_d(option = "magma")
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Greenhouse gas emissions are divided into sectors, as seen in the graph above. Except for Iceland, it appears that each country have similar proportions in which sectors are contributing the most and least to their overall emissions. Energy is the highest for all (again, except for Iceland), with transport second. Agriculture and waste are the lowest, although the US has it ranking in the middle instead. Land use refers to carbon emissions from the general land - this also includes forests, which can absorb carbon. This results in what is termed carbon sequestration[^4]. As it turns out, Monaco, Japan, the Russian Federation, and the USA all have negative emissions from land use, or carbon sequestration, so this has been excluded from the proportions, as they are not contributing to the country's overall emissions [^5] [^6]. Strangely, Iceland is an exception, with land use emissions dominating in its total emissions. This seems to be due to the relatively low amount of forest land it has compared to other categories of land that produce positive emission. However, there has been discussion about wetland restoration efforts that, while in the short run may increase methane emissions, can also in the long run result in overall cuts in emissions. This may be something worth continued pursuit from Iceland.[^7]

[^4]: tdus. “Biological Carbon Sequestration.” UC Davis, 21 Aug. 2021, <https://www.ucdavis.edu/climate/definitions/carbon-sequestration/biological>.

[^5]: US EPA, OAR. "Sources of Greenhouse Gas Emissions." 29 Dec. 2015, <https://www.epa.gov/ghgemissions/sources-greenhouse-gas-emissions>.

[^6]: "Land Use, Land-Use Change and Forestry (LULUCF)." United Nations Framework Convention on Climate Change, United Nations, <https://unfccc.int/topics/land-use/workstreams/land-use--land-use-change-and-forestry-lulucf.> Accessed 10 Dec. 2024.

[^7]: island.is. "Land and forest iceland. Land use emissions rise, but binding in forests rises." 5 Jun. 2024, <https://island.is/en/o/land-and-forest-iceland/news/losun-vegna-landnotkunar-eykst-en-binding-eykst-i-skogum>.


```{r GHG Emission Levels by Sector, fig.cap = "Figure 5. The US has the highest greenhouse gas emissions across all sectors except for land use, where Iceland leads instead. Except for land use, the US is followed by the Russian Federation, Japan, Iceland, Malta, and lastly Monaco. Data was compiled by the United Nations from countries around the world according to the latest year for recorded data available. Greenhouse gas emissions are measured in orders of mangitude, or powers of 10, of tonnes of CO2 equivalent.", fig.alt = "This is a faceted bar chart with the countries Monaco, Malta, Iceland, Japan, the Russian Federation, and the USA on the x-axis. The y-axis is greenhouse gas emissions in orders of magnitude, from 0 to a little under 10 orders of magnitude, or ten billion. This data is also faceted into six copies, each one for a different sector: agriculture, energy, transport, industrial, land use, and waste. The US has the highest greenhouse gas emissions across all sectors except for land use, where Iceland leads instead. Except for land use, the US is followed by the Russian Federation, Japan, Iceland, Malta, and lastly Monaco."}
tidy_sector |> 
  # Convert from thousands to ones, then calculate orders of magnitude
  mutate(GHG_Emissions_magnitude = log10(GHG_Emissions_thou * 1000)) |> 
  ggplot() +
  geom_col(aes(x = Country,
               y = GHG_Emissions_magnitude,
               fill = Country),
           show.legend = F) +
  facet_wrap(~Sector) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  # Change y-axis labels to clarify the meaning of orders of magnitude
  scale_y_continuous(position = "left",
                     breaks = c(0, 2, 4, 6, 8, 10),
                     labels = parse(text = c("0", "10^2", "10^4",
                                             "10^6", "10^8", "10^10")),
                     # Duplicate labels to right side of graph
                     sec.axis = dup_axis(name = "")) +
  labs(title = "Greenhouse Gas Emissions by Sector",
       subtitle = "2021",
       y = "GHG Emissions in Orders of Magnitude") +
  theme_dark() +
  scale_fill_viridis_d(option = "magma")
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;On the flip side of the coin from the previous graph, here we see the actual numbers of emission levels from each sector for each country. We can observe a very similar pattern here as we did from the CO2 emission levels graphs, where the US leads in all sectors except for land use, followed by the Russian Federation, Japan, Iceland, Malta, and lastly Monaco.

## Conclusion

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;It seems clear based on our explorations that the United States of America is the one of the highest greenhouse gas producers in the world, even going by sector or taking population into account. This suggests that when constructing equitable climate policies around the world, we may have to be stricter and harsher with the USA for the country to reach the same lower levels as some other countries. Additionally, we observe that countries like Iceland and the Russian Federation have had gradual increases in overall carbon dioxide emissions over the past couple decades - thus it may be imperative to identify the reasons for these increases, that are not observed in the other countries including the USA, and how to stop that unwanted growth. When tackling each country's emissions individually, it seems a reasonable priority would be to understand what makes the energy sector such a large emitter, and invest more into pursuing solutions for reducing energy emissions. The exception in the countries of our focus is Iceland, which instead has land use dominating. This is a sharp contrast from most of the other countries which had negative land use emissions - understanding the reasoning for this and figuring out what we could do about it may also be beneficial. Overall, these data visualizations help us to get an idea of where potential problems may lie and inform us in the best way forward with constructing climate policies.

## Works Cited